### 2019-05-09 2-Chapters/4-Chapter/IEEE_WCNC_2019__DemoICT.git/main.tex
I'm happy about the modifications for Chapter 5:
- no proof for lower-bound
- less figures
- shorter Appendix
- few more details in a proof
- less clutter in some equations

\paragraph{Related work.}
% FIXME
\TODOL{Est-ce que je fais ça ici ou à la fin du chapitre 5 ? Ou peut-être en chapitre 1, ça ferait du contenu au chapitre 1...}
We evoke quickly here some previous works that similarly implemented proof-of-concepts to show that reinforcement learning algorithms can be used within real-world wireless communications.
\cite{darak2016bayesian}
\cite{Darak16}
\cite{modiDemo2016}
\cite{kumar2016two}
\cite{KumarYadav2018}
\cite{SawantKumar2018}
\cite{JoshiKumar2018}

@@

\begin{figure}[!b]
    \centering
    \includegraphics[width=0.70\linewidth]{SCEE_TestBed1.jpg}
    \vspace*{20pt}
    \includegraphics[width=0.70\linewidth]{SCEE_TestBed2.jpg}
    \caption{Two pictures showing the SCEE testbed \cite[Appendix~3]{Bodinier17}, taken in $2018$.}
    \label{fig:42:photosSCEETestBed}
\end{figure}

>>>>

\paragraph{Related works.}
% FIXME?
% \TODOL{Est-ce que je fais ça ici ou à la fin du chapitre 5 ? Ou peut-être en chapitre 1, ça ferait du contenu au chapitre 1...}
%
We evoke quickly here some previous works that similarly implemented proof-of-concepts to show that reinforcement learning algorithms can be used within real-world wireless communications.
In $2015$ and $2016$, a PhD student (Navikkumar Modi) and post-doc (Sumit J. Darak) in our team SCEE worked with Christophe Moy and others for the following publications,
\cite{darak2016bayesian,Darak16,modiDemo2016,kumar2016two}.
Since $2017$, Sumit J. Darak and his team at IIIT Delhi have been very active in the research on cognitive radio using multi-armed bandits, and some of their recent works are illustrated with real-world demo using USRP and the MATLAB/Simulink system
\cite{KumarYadav2018,SawantKumar2018,JoshiKumar2018}.

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.70\linewidth]{SCEE_TestBed1.jpg}
    \vspace*{20pt}
    \includegraphics[width=0.70\linewidth]{SCEE_TestBed2.jpg}
    \caption{Two pictures showing the SCEE testbed \cite[Appendix~3]{Bodinier17}, taken in $2018$.}
    \label{fig:42:photosSCEETestBed}
\end{figure}

### 2019-05-09 2-Chapters/4-Chapter/IEEE_WCNC_2019__DemoICT.git/main_appendix.tex
HARD
I'm happy about the modifications for Chapter 5:
- no proof for lower-bound
- less figures
- shorter Appendix
- few more details in a proof
- less clutter in some equations

For the curious reader, we wanted to include here an illustration and a description of the three components of our demonstration, presented above in Section~\ref{sec:4:gnuradio}.

@@

\begin{figure}[!h]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_TX_PU__v1__simple_grc.png}
    \caption{The \textbf{random traffic generator} flow-graph.}
    \label{fig:4app:USRP_TX_PU__v1__simple_grc}
\end{figure}

The \textbf{IoT gateway} is presented in Figure~\ref{fig:4app:USRP_TX_PU__v1__simple_grc}.
%
The hand-written blocks are the 1) \texttt{Demodulator} block, which is configured with a list of detection threshold (on the received power), 2) the \texttt{Check\_ack} block, which is configured with a maximum block error and an accepted error rates (to decide when a message is close enough to an acknowledgement), and finally 3) the \texttt{send\_ack} block, which is configured with the list of active channels, and knowledge about the PHY layer (delay before sending the Ack and its length). All blocks need to know constants about the PHY layer (preamble length and data length). The IoT gateway uses one USRP for both emitting and receiving data (``Sink'' and ``Source'' modes).

\begin{figure}[!h]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_RX_BTS__v1__simple_grc.png}
    \caption{The \textbf{IoT gateway} flow-graph.}
    \label{fig:4app:USRP_RX_BTS__v1__simple_grc}
\end{figure}

Figure~\ref{fig:4app:USRP_TX_SU__v1__simple_grc} below shows the \textbf{IoT dynamic device} flow-graph.
The hand-written blocks are the 1) the \texttt{Renormalized\_ack} block, which is configured with the list of active channels and extra knowledge about the PHY layer (preamble length, message length, and a threshold to tune detection of the incoming Ack), 2) the \texttt{Demodulator} and 3) the \texttt{Check\_ack} blocks, both shared with the IoT gateway, and finally 4) the \texttt{generator\_SU} block, which embeds the \UCB{} or Thompson Sampling algorithm, and which is configured with the list of active channels. Most blocks need to know constants about the PHY layer (preamble length and data length). Any of the IoT dynamic devices use one USRP for both emitting and receiving data (``Sink'' and ``Source'' modes)

\begin{figure}[!h]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_TX_SU__v1__simple_grc.png}
    \caption{The \textbf{IoT dynamic device} flow-graph.}
    \label{fig:4app:USRP_TX_SU__v1__simple_grc}
\end{figure}

>>>>

For the curious reader, we wanted to include here an illustration and a description of the three components of the demonstration we presented above in Section~\ref{sec:4:gnuradio}.
%
The code corresponding to the following components is written in \texttt{C++} and \texttt{Python} for the hand-written blocks, and using the GUI of the GNU Radio Companion software.
The flow-graphs are saved as XML files, and the complete code of our demo is fully available online, open-sourced under GPLv3 license, at
\href{https://bitbucket.org/scee_ietr/malin-multi-arm-bandit-learning-for-iot-networks-with-grc}{\texttt{bitbucket.org/scee\_ietr/malin-multi-arm-bandit-learning}\\\texttt{-for-iot-networks-with-grc/}}.

\subsubsection*{Random traffic generator}

@@

\begin{figure}[!h]
	% \centering
    \includegraphics[width=1.02\textwidth]{2-Chapters/4-Chapter/Images/USRP_TX_PU__v1__simple_grc.png}
    \caption{The \textbf{random traffic generator} flow-graph.}
    \label{fig:4app:USRP_TX_PU__v1__simple_grc}
\end{figure}

\subsubsection*{IoT gateway}

The \textbf{IoT gateway} is presented in Figure~\ref{fig:4app:USRP_RX_BTS__v1__simple_grc} below.
%
We wrote the following blocks:
\begin{itemize}
    \item
    1) the \texttt{Demodulator} block, which is configured with a list of detection threshold (on the received power),
    \item
    2) the \texttt{Check\_ack} block, which is configured with a maximum block error and an accepted error rates (to decide when a message is close enough to an acknowledgement),
    \item
    and finally
    3) the \texttt{send\_ack} block, which is configured with the list of active channels, and knowledge about the PHY layer (delay before sending the Ack and its length).
\end{itemize}
All blocks need to know constants about the PHY layer (preamble length and data length). The IoT gateway uses one USRP for both emitting and receiving data (``Sink'' and ``Source'' modes).

\begin{figure}[!h]
	% \centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_RX_BTS__v1__simple_grc.png}
    \caption{The \textbf{IoT gateway} flow-graph.}
    \label{fig:4app:USRP_RX_BTS__v1__simple_grc}
\end{figure}

\subsubsection*{IoT dynamic device}

Figure~\ref{fig:4app:USRP_TX_SU__v1__simple_grc} below shows the \textbf{IoT dynamic device} flow-graph.
The hand-written blocks are the
\begin{itemize}
    \item
    1) the \texttt{Renormalized\_ack} block, which is configured with the list of active channels and extra knowledge about the PHY layer (preamble length, message length, and a threshold to tune detection of the incoming Ack),
    \item
    2) the \texttt{Demodulator} and 3) the \texttt{Check\_ack} blocks, both shared with the IoT gateway,
    \item
    and finally 4) the \texttt{generator\_SU} block, which embeds the \UCB{} or Thompson Sampling algorithm, and which is configured with the list of active channels.
\end{itemize}
Most blocks need to know constants about the PHY layer (preamble length and data length). Any of the IoT dynamic devices use one USRP for both emitting and receiving data (``Sink'' and ``Source'' modes)

\begin{figure}[!h]
	% \centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_TX_SU__v1__simple_grc.png}
    \caption{The \textbf{IoT dynamic device} flow-graph.}
    \label{fig:4app:USRP_TX_SU__v1__simple_grc}
\end{figure}

### 2019-05-09 2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM.git/IEEE_WCNC__2019__Paper__BMBM.tex
Talk to Emilie
I'm happy about the modifications for Chapter 5:
- no proof for lower-bound
- less figures
- shorter Appendix
- few more details in a proof
- less clutter in some equations

\begin{figure}[htp!]  % [htbp]
	\centering
	% \includegraphics[width=0.65\linewidth]{Approximation_m10.eps}
	\includegraphics[width=0.65\linewidth]{Approximation_m10-eps-converted-to.pdf}
	\caption[Proposed approximation for the probability of collision at the second transmission]{Proposed approximation for the probability of collision at the second transmission. It is more precise for smaller values of $N$.}
	\label{fig:43:Approximation_m10}
\end{figure}

@@

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$\;
				Transmit in channel $A(t) = \arg\max_{1\leq k \leq K} U_k(t)$\;
				Reward $r_{A(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
			\Else(\tcp*[f]{Retransmit in same channel}){
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Transmit in channel $A(t) = j$\;
			}
		}
		\caption{Heuristic: first-stage \UCB{} and retransmission in same channel.}
		\label{algo:43:UCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\paragraph{A warning about notations: local (device) time vs global (gateway) time.}
%
In all the algorithms and notations presented in this Section, the time steps $t$ do not refer to the global time steps (as seen from the gateway), but rather the current number of uplink transmissions (or retransmissions) that was carried out by the device.
In other words, each device wakes up whenever it has to send some data, following its random emission process (here, we propose a Bernoulli process of small probability, \eg, $p=10^{-3}$), and when it woke up, it sends its packet, and will send again for at most $M$ times (\eg, $M=10$) until it receives an \Ack.
To be clear, once again, we repeat that the time steps $t$ in the following algorithms denote the number of time the device sent an uplink packet, and these uplink (re)transmissions can happen at different (real-world) times, more or less spread out in time, it does not matter to the learning part of this model and work.

@@

\vspace*{-3pt}
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
	% \begin{small} % XXX remove if needed
	\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Random retransmission}){
				Transmit in channel $A(t) \sim \mathcal{U}(1,\ldots,K)$\;
			}
		}
		\caption{Heuristic: uniform random retransmission.}    % A naive
		\label{algo:43:UCBthenRandom}
	% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^r$}){
				Compute $\forall k, U^r_k(t) = \widehat{\mu^r_k}(t) + \sqrt{\alpha \log(t) / N_k^r(t)}$\;
				Transmit in channel $C^r(t) = \arg\max_{1\leq k \leq K} U^r_k(t)$\;
				Reward $r^r_{C^r(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: \UCB{} for retransmission.}    % A naive
		\label{algo:43:TwoUCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		% \KwData{$\forall k,j\in[\![1;K]\!]$, $N_k^j(t)$, $\widehat{\mu_k}^j(t)$, $B_k^j(t)$ and $U_k^j(t)$}
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^j$}){ % Retr. after trying channel $j$
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Compute $\forall k, U_k^j(t) = \widehat{\mu_k}^j(t) + \sqrt{\alpha \log(t) / N_k^j(t)}$\;
				Transmit in channel $A^j(t) = \arg\max_{1\leq k \leq K} U^j_k(t)$\;
				Reward $r^j_{A^j(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: $K$ different {\UCB}s for retransmission.}
		\label{algo:43:UCBthenKp1}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

The idea behind this delay is to allow the first stage \UCB{} to start learning the best channel, before starting the second stage \UCB{} (see details in Algorithm~\ref{algo:43:UCBwithDelay}).
The number of transmissions to wait before applying the second algorithm is denoted by $\Delta$, it has to be fixed before-hand\footnote{~Choosing the value of $\Delta$ could be done by extensive benchmarks but such approach goes against the reinforcement learning idea: an heuristic should work against any problem, without the need to simulate the problem before-hand to find a good value of some internal parameter. As such, we only consider a delay of $\Delta=100$.}.
%
Note that, we use the subscript $({}^d)$ to denote the variables
related to the delayed second-stage \UCB{} algorithm.

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
	\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\uElseIf(\tcp*[f]{Random retransmission}){$t \leq \Delta$}{
				Transmit randomly in a channel $A(t) \sim \mathcal{U}(1,\ldots,K)$.
			}
			\Else(\tcp*[f]{Packet retransmission with delayed $\mathrm{UCB}^d$}){
				Compute $\forall k, U^d_k(t) = \widehat{\mu^d_k}(t) + \sqrt{\alpha \log(t) / N_k^d(t)}$\;
				Transmit in channel $A^d(t) = \arg\max_{1\leq k \leq K} U^d_k(t)$\;
				Reward $r^d_{A^d(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: Delayed \UCB{} for retransmission.}
		\label{algo:43:UCBwithDelay}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\paragraph{Other ideas that we did not explore}
%
Instead of considering a first-stage \UCB{} followed by a random channel retransmission, we could have considered a random first-stage channel retransmission followed by a second-stage \UCB.
This was not considered in our study \cite{Bonnefoi2019WCNC}, and we preferred to not add it to the experiments presented below, for three reasons: to avoid clutter in the plots, to win some time as these experiments had to run for quite a long time, but mainly because the first-stage channel selection is most important one as we illustrate by the large difference in terms of performance between the uniform channel selection and the ``only UCB'' heuristic, below in Figures~\ref{fig:43:mainExperiment1} and \ref{fig:43:mainExperiment2}.

\textbf{Remark.}
%
Similarly, Philippe Mary asked me at the MoTION workshop where I presented \cite{Bonnefoi2019WCNC}, why didn't we rather used Bayes-UCB instead of \UCB{} (from \cite{Kaufmann12BUCB}), as it is known to empirically outperform \UCB{} (as illustrated for instance in \cite{darak2016bayesian,kumar2016two}).
We could have chose any bandit algorithm for the ``first stage'' component used in this Section, and for simplicity and clarity we focussed on a simple but efficient one (\UCB).
We believe that the same empirical results, but more importantly, the same conclusions, could be given if we used Bayes-UCB or other bandit algorithm for the base building block in the different heuristics proposed in Section~\ref{sub:43:heuristics}.

>>>>

\begin{figure}[htp!]  % [htbp]
	\centering
	% \includegraphics[width=0.65\linewidth]{Approximation_m10.eps}
	\includegraphics[width=0.80\linewidth]{Approximation_m10-eps-converted-to.pdf}
	\caption[Proposed approximation for the probability of collision at the second transmission]{Proposed approximation for the probability of collision at the second transmission. It is more precise for smaller values of $N$.}
	\label{fig:43:Approximation_m10}
\end{figure}

@@

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$\;
				Transmit in channel $A(t) = \arg\max_{1\leq k \leq K} U_k(t)$\;
				Reward $r_{A(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
			\Else(\tcp*[f]{Retransmit in same channel}){
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Transmit in channel $A(t) = j$\;
			}
		}
		\caption{First-stage \UCB{} and retransmission in same channel.}
		\label{algo:43:UCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\paragraph{\textcolor{red}{Small warning about notations: local (device) time vs global (gateway) time}.}
%
In all the algorithms and notations presented in this Section, the time steps $t$ do not refer to the global time steps (as seen from the gateway), but rather the current number of uplink transmissions (or retransmissions) that was carried out by the device.
In other words, each device wakes up whenever it has to send some data, following its random emission process (here, we propose a Bernoulli process of small probability, \eg, $p=10^{-3}$), and when it woke up, it sends its packet, and will send again for at most $M$ times (\eg, $M=10$) until it receives an \Ack.

To be clear, once again, we repeat that the time steps $t$ in the following algorithms denote the number of time the device sent an uplink packet, and these uplink (re)transmissions can happen at different (real-world) times, more or less spread out in time, it does not matter to the learning part of this model and work.

@@

% \vspace*{-3pt}
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
	% \begin{small} % XXX remove if needed
	\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Random retransmission}){
				Transmit in channel $A(t) \sim \mathcal{U}(1,\ldots,K)$\;
			}
		}
		\caption{Uniform random retransmission.}    % A naive
		\label{algo:43:UCBthenRandom}
	% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}

@@

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^r$}){
				Compute $\forall k, U^r_k(t) = \widehat{\mu^r_k}(t) + \sqrt{\alpha \log(t) / N_k^r(t)}$\;
				Transmit in channel $C^r(t) = \arg\max_{1\leq k \leq K} U^r_k(t)$\;
				Reward $r^r_{C^r(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{\UCB{} for retransmission.}    % A naive
		\label{algo:43:TwoUCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		% \KwData{$\forall k,j\in[\![1;K]\!]$, $N_k^j(t)$, $\widehat{\mu_k}^j(t)$, $B_k^j(t)$ and $U_k^j(t)$}
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^j$}){ % Retr. after trying channel $j$
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Compute $\forall k, U_k^j(t) = \widehat{\mu_k}^j(t) + \sqrt{\alpha \log(t) / N_k^j(t)}$\;
				Transmit in channel $A^j(t) = \arg\max_{1\leq k \leq K} U^j_k(t)$\;
				Reward $r^j_{A^j(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{$K$ different {\UCB}s for retransmission.}
		\label{algo:43:UCBthenKp1}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

The idea behind this delay is to allow the first stage \UCB{} to start learning the best channel, before starting the second stage \UCB{} (see details in Algorithm~\ref{algo:43:UCBwithDelay}).
The number of transmissions to wait before applying the second algorithm is denoted by $\Delta$, it has to be fixed before-hand\footnote{~Choosing the value of $\Delta$ could be done by extensive benchmarks but such approach goes against the reinforcement learning idea: an heuristic should work against any problem, without the need to simulate the problem before-hand to find a good value of some internal parameter. As such, we only consider a delay of $\Delta=100$ in our experiments, and we did not try to optimize it.}.
%
Note that, we use the subscript $({}^d)$ to denote the variables
related to the delayed second-stage \UCB{} algorithm.

% \vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
	\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\uElseIf(\tcp*[f]{Random retransmission}){$t \leq \Delta$}{
				Transmit randomly in a channel $A(t) \sim \mathcal{U}(1,\ldots,K)$.
			}
			\Else(\tcp*[f]{Packet retransmission with delayed $\mathrm{UCB}^d$}){
				Compute $\forall k, U^d_k(t) = \widehat{\mu^d_k}(t) + \sqrt{\alpha \log(t) / N_k^d(t)}$\;
				Transmit in channel $A^d(t) = \arg\max_{1\leq k \leq K} U^d_k(t)$\;
				Reward $r^d_{A^d(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Delayed \UCB{} for retransmission.}
		\label{algo:43:UCBwithDelay}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\paragraph{Other ideas that we did not explore.}
%
Instead of considering a first-stage \UCB{} followed by a random channel retransmission, we could have considered a random first-stage channel retransmission followed by a second-stage \UCB.
This was not considered in our study \cite{Bonnefoi2019WCNC}, and we preferred to not add it to the experiments presented below, for three reasons: to avoid clutter in the plots, to win some time as these experiments had to run for quite a long time, but mainly because the first-stage channel selection is most important one as we illustrate by the large difference in terms of performance between the uniform channel selection and the ``only UCB'' heuristic, below in Figures~\ref{fig:43:mainExperiment1} and \ref{fig:43:mainExperiment2}.

\paragraph{Using another bandit policy?}
%
Similarly, Philippe Mary asked me at the MoTION workshop where I presented \cite{Bonnefoi2019WCNC}, why didn't we rather used Bayes-UCB instead of \UCB{} (from \cite{Kaufmann12BUCB}), as it is known to empirically outperform \UCB{} (as illustrated for instance in \cite{darak2016bayesian,kumar2016two}).
We could have chose any bandit algorithm for the ``first stage'' component used in this Section, and for simplicity and clarity we focussed on a simple but efficient one (\UCB).
We believe that the same empirical results, but more importantly, the same conclusions, could be given if we used Bayes-UCB or other bandit algorithm for the base building block in the different heuristics proposed in Section~\ref{sub:43:heuristics}.
The goal of this Section was not to optimize on the bandit policy (as we presented in Section~\ref{sec:3:reviewSPAlgorithms} some numerical simulations doing precisely this), but rather to compare the different heuristics, for a fixed bandit policy (for which we preferred to use \UCB{} for simplicity).

### 2019-05-09 2-Chapters/4-Chapter/chapter4.tex
I'm happy about the modifications for Chapter 5:
- no proof for lower-bound
- less figures
- shorter Appendix
- few more details in a proof
- less clutter in some equations

% ----------------------------------------------------------------------------
\section{Conclusion}
\label{sec:4:conclusion}
% ----------------------------------------------------------------------

We give in this section conclusions about the two models presented above, one for long-life IoT devices (without retransmission) and for improved QoS devices (\ie, with retransmissions), and about our demonstration, as well as many directions of future works, some of which are currently being studied by researchers of our team.

\TODOL{Je devrais réécrire toute cette conclusion, si j'ai le temps, pour faire mieux qu'une suite de 3 conclusions et 3 future works, et être plus concis et moins ``short sighted'' comme disait Emilie.}

\subsection*{Conclusions}

@@

In Section~\ref{sec:4:firstModel}, we proposed an evaluation of the performance of MAB learning algorithms in IoT networks,
with a focus on the convergence of algorithms, in terms of successful transmission rates, when the proportion of intelligent dynamic devices changes.
Concretely, increasing this probability allows to insert more devices in the same network, while maintaining a good Quality of Service.
This approach can also extend IoT devices battery life, as they suffer less from failed transmissions.
We show that \UCB{} and TS have near-optimal performance, even when their underlying \iid{} assumption is violated by the presences many ``intelligent'' end-devices which follow a random activation process.

This is both a surprising and a very encouraging result, showing that application of bandit algorithms tailored for a stochastic model is still useful in broader settings.
The fully \emph{decentralized} application of classic stochastic MAB algorithms are almost as efficient as the best possible centralized policy in this setting, after a short learning period, even though the dynamic devices \emph{can not} communicate with each others, and \emph{do not} know the system parameters.

@@

In Section~\ref{sec:4:retransmissions}, we presented an extension of our model of LPWA networks based on an ALOHA protocol, slotted both in time and frequency, in which dynamic IoT devices can again use machine learning algorithms, to improve their Packet Loss Ratio (PLR) when accessing the network.
If the priority is the Quality of Service (QoS), for instance with renewable energy capabilities, this second model is more interesting.
The main novelty of this model is to address the packet retransmissions upon radio collision, by using a Multi-Armed Bandit framework.
We presented and evaluated several learning heuristic that try to learn how to transmit and retransmit in a smarter way, by using the \UCB{} algorithm for channel selection for first transmission, and different proposals based on \UCB{} for the retransmissions upon collisions.

We showed that incorporating learning for the transmission is needed to achieve optimal performance, with significant gain in terms of successful transmission rate in networks with a large number of devices (up-to $30\%$ in the example network).
Our empirical simulations show that each of our proposed heuristic outperforms a naive random access scheme.
Surprisingly, the main take-away message is that a simple \UCB{} learning approach, that retransmit in the same channel, turns out to perform as well as more complicated heuristics.

@@

Future works related to this chapter include the following directions.

>>>>

% \newpage  % WARNING ?
% ----------------------------------------------------------------------------
\section{Conclusion}
\label{sec:4:conclusion}
% ----------------------------------------------------------------------

% We give in this section conclusions about the two models presented above, one for long-life IoT devices (without retransmission) and for improved QoS devices (\ie, with retransmissions), and about our demonstration, as well as directions of future works, some of which are currently being studied by researchers of our team.

% \TODOL{Je devrais réécrire toute cette conclusion, si j'ai le temps, pour faire mieux qu'une suite de 3 conclusions et 3 future works, et être plus concis et moins ``short sighted'' comme disait Emilie.}

% \subsection*{Conclusions}

@@

In Section~\ref{sec:4:firstModel}, we proposed an evaluation of the performance of MAB learning algorithms in IoT networks,
with a focus on the convergence of algorithms, in terms of successful transmission rates, when the proportion of intelligent dynamic devices changes.
Concretely, increasing this probability allows to insert more devices in the same network, while maintaining a good Quality of Service.
Similarly, if the number of devices remain constant, increasing the successful transmission rate directly extends the IoT devices battery life, as they suffer less from failed transmissions.
We show that \UCB{} and TS have near-optimal performance, even when their underlying \iid{} assumption (see Chapter~\ref{chapter:2}) is violated by the presence of many ``intelligent'' end-devices which follow a random activation process.
%
This is both surprising and encouraging, it shows that applying bandit algorithms tailored for a stochastic model is still useful in broader settings.
The fully \emph{decentralized} application of classic stochastic MAB algorithms are almost as efficient as the best possible centralized policy in this setting, after a short learning period, even though the dynamic devices \emph{cannot} communicate with each others, and \emph{do not} know the system parameters.

@@

In Section~\ref{sec:4:retransmissions}, we presented an extension of our model of LPWA networks based on an ALOHA protocol, slotted both in time and frequency.
% , in which dynamic IoT devices can again use machine learning algorithms, to improve their Packet Loss Ratio (PLR) when accessing the network.
If the priority is the Quality of Service (QoS), for instance with renewable energy capabilities, this second model is more appropriate.
The main novelty of this model is to address the packet retransmissions upon radio collision, by using a Multi-Armed Bandit framework.
We presented and evaluated several heuristics that try to learn how to transmit and retransmit in a smarter way, by using the \UCB{} algorithm for channel selection for first transmission, and different proposals based on \UCB{} for the retransmissions upon collisions.
%
We showed that incorporating learning for the transmission is needed to achieve optimal performance, with significant gain in terms of successful transmission rate in networks with a large number of devices (up-to $30\%$ in the example network).
Our empirical simulations show that each of our proposed heuristic outperforms a naive random access scheme.
Surprisingly, the main take-away message is that a simple \UCB{} learning approach, that retransmit in the same channel, turns out to perform as well as more complicated heuristics.

@@

% Future works related to this chapter include the following directions.
